# Some EDA

# NY state crime data, overview:

```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(patchwork)
library(gifski)
library(transformr)
library(dplyr)
library(ggplot2)
library(forcats)
library(gganimate)
```

## Data Checks:

```{r}

crime_ny <- read.csv("../data/Index_Crimes_by_County_and_Agency__Beginning_1990.csv")
crime_nyc <- read.csv('../data/NYPD_Shooting_Incident_Data__Historic__20231124.csv')
pop_ny <- read.csv('../data/Annual_Population_Estimates_for_New_York_State_and_Counties__Beginning_1970.csv')
pop_ny$FIPS.Code <- substr(pop_ny$FIPS.Code, 3, nchar(pop_ny$FIPS.Code))

county_fips <- data.frame(
  FIPS = c(
    "001", "003", "005", "007", "009", "011", "013", "015", "017", "019", "021",
    "023", "025", "027", "029", "031", "033", "035", "037", "039", "041",
    "043", "045", "047", "049", "051", "053", "055", "057", "059", "061",
    "063", "065", "067", "069", "071", "073", "075", "077", "079", "081",
    "083", "085", "087", "089", "089", "091", "093", "095", "097", "099", "101",
    "103", "105", "107", "109", "111", "113", "115", "117", "119", "121", "123"
  ),
  Name = c(
    "Albany", "Allegany", "Bronx", "Broome", "Cattaraugus", "Cayuga", "Chautauqua",
    "Chemung", "Chenango", "Clinton", "Columbia", "Cortland", "Delaware",
    "Dutchess", "Erie", "Essex", "Franklin", "Fulton", "Genesee", "Greene",
    "Hamilton", "Herkimer", "Jefferson", "Kings", "Lewis", "Livingston",
    "Madison", "Monroe", "Montgomery", "Nassau", "New York", "Niagara",
    "Oneida", "Onondaga", "Ontario", "Orange", "Orleans", "Oswego",
    "Otsego", "Putnam", "Queens", "Rensselaer", "Richmond", "Rockland",
    "St. Lawrence", "St Lawrence", "Saratoga", "Schenectady", "Schoharie", "Schuyler",
    "Seneca", "Steuben", "Suffolk", "Sullivan", "Tioga", "Tompkins",
    "Ulster", "Warren", "Washington", "Wayne", "Westchester", "Wyoming", "Yates"
  )
)
crime_ny$County <- as.character(crime_ny$County)
crime_ny_with_fips <- merge(crime_ny, county_fips, by.x = "County", by.y = "Name", all.x = TRUE)
crime_ny_with_fips <- crime_ny_with_fips[crime_ny_with_fips$Agency == 'County Total', ]

crime_ny_with_fips_and_pop <- merge(crime_ny_with_fips, 
                                    pop_ny, by.x = c("FIPS", "Year"), 
                                    by.y = c("FIPS.Code", "Year"), 
                                    all.x = TRUE)
crime_ny_with_fips_and_pop <- crime_ny_with_fips_and_pop[!is.na(crime_ny_with_fips_and_pop$Index.Total), ]
crime_ny_with_fips_and_pop <- crime_ny_with_fips_and_pop[, !(names(crime_ny_with_fips_and_pop) %in% c("Geography", "Program.Type"))]

na_cols <- colSums(is.na(crime_ny_with_fips_and_pop))
cat("Columns with NA values:", names(na_cols[na_cols>0]), "\n")
na_rows <- crime_ny_with_fips_and_pop[is.na(crime_ny_with_fips_and_pop[, names(na_cols[na_cols>0])]), ]
cat("Rows with NA values:", nrow(subset(na_rows, !(Agency == "County Total"))), "\n")

na_cols <- colSums(is.na(crime_ny))
cat("Columns with NA values:", names(na_cols[na_cols>0]), "\n")

na_rows <- crime_ny[is.na(crime_ny[, names(na_cols[na_cols>0])]), ]

cat("Rows with NA values:", nrow(subset(na_rows, !(Agency == "County Total"))), "\n")
cat("Newest with NA values:",max((subset(na_rows, !(Agency == "County Total")))$Year), "\n")

head(crime_nyc,5)

```






# Population dataset:

```{r}
pop_ny <- read.csv('../data/Annual_Population_Estimates_for_New_York_State_and_Counties__Beginning_1970.csv')
pop_ny$FIPS.Code <- substr(pop_ny$FIPS.Code, 3, nchar(pop_ny$FIPS.Code))

na_cols <- colSums(is.na(pop_ny))
cat("Columns with NA values:", names(na_cols[na_cols>0]), "\n")

```




```{r}

pop_ny_full_state <- pop_ny[pop_ny$Geography == 'New York State', ]

population_plot <- ggplot(pop_ny_full_state, 
                          aes(x = Year, 
                              y = Population, 
                              fill = Program.Type)) +
  geom_col(position = "dodge") +
  labs(title = "Population per Year", 
       x = "Year", 
       y = "Population") +
  scale_fill_manual(values = c("Census Base Population" = "blue", 
                              "Postcensal Population Estimate" = "red",
                              'Intercensal Population Estimate' = 'orange')) +
  theme_minimal()

plot(population_plot)

```


```{r}
library(tidyr)


census_years <- unique(pop_ny_full_state[pop_ny_full_state$Program.Type == 'Census Base Population', ]$Year)

pop_ny_full_state_cens_yrs <- subset(pop_ny_full_state, pop_ny_full_state$Year %in% census_years)

pop_ny_full_state_cens_yrs_wide <- pivot_wider(pop_ny_full_state_cens_yrs, 
                       names_from = Program.Type, 
                       values_from = Population)
colnames(pop_ny_full_state_cens_yrs_wide)

names(pop_ny_full_state_cens_yrs_wide)

ggplot(pop_ny_full_state_cens_yrs_wide, aes(x = Year)) +
  geom_point(aes(y = `Census Base Population`, color = "Census Base Population"), 
             position = position_dodge(width = 0.4)) +
  geom_point(aes(y = `Intercensal Population Estimate`, color = "Intercensal Estimate"), 
             position = position_dodge(width = 0.2)) +
  geom_point(aes(y = `Postcensal Population Estimate`, color = "Postcensal Estimate"), 
             position = position_dodge(width = -0.2)) +
  labs(title = "A", x = "B", y = "C") +
  scale_color_manual(values = c("Census Base Population" = "blue", 
                                "Intercensal Estimate" = "orange",
                                "Postcensal Estimate" = "red"),
                     name = "Legend Title") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```





```{r}

# Find mismatch in County Total vs per-police station:

county_total_entries <- crime_ny %>%
  filter(Agency == "County Total")

columns_to_sum <- c(
  "Index.Total", "Violent.Total", "Murder", "Rape",
  "Robbery", "Aggravated.Assault", "Property.Total",
  "Burglary", "Larceny", "Motor.Vehicle.Theft"
)

crime_ny_not_ct <- crime_ny[crime_ny$Agency!='County Total',] 

summed_data <- crime_ny_not_ct %>%
  group_by(Year, County) %>%
  summarise(across(all_of(columns_to_sum), sum), .groups = "drop")

comparison_result <- left_join(summed_data, 
                               county_total_entries, 
                               by = c("Year", "County"))

comparison_result <- comparison_result %>%
  mutate(
    Check_Index.Total = ifelse(Index.Total.x == Index.Total.y, "Match", "Mismatch"),
    Check_Violent.Total = ifelse(Violent.Total.x == Violent.Total.y, "Match", "Mismatch"),
    Check_Murder = ifelse(Murder.x == Murder.y, "Match", "Mismatch"),
    Check_Rape = ifelse(Rape.x == Rape.y, "Match", "Mismatch"),
    Check_Robbery = ifelse(Robbery.x == Robbery.y, "Match", "Mismatch"),
    Check_Aggravated.Assault = ifelse(Aggravated.Assault.x == Aggravated.Assault.y, "Match", "Mismatch"),
    Check_Property.Total = ifelse(Property.Total.x == Property.Total.y, "Match", "Mismatch"),
    Check_Burglary = ifelse(Burglary.x == Burglary.y, "Match", "Mismatch"),
    Check_Larceny = ifelse(Larceny.x == Larceny.y, "Match", "Mismatch"),
    Check_Motor.Vehicle.Theft = ifelse(Motor.Vehicle.Theft.x == Motor.Vehicle.Theft.y, "Match", "Mismatch")
  )

check_columns <- paste0("Check_", columns_to_sum)
mismatch_rows <- comparison_result %>%
  rowwise() %>%
  filter(any(c_across(all_of(check_columns)) == "Mismatch"))

mismatch_rows_long <- tidyr::gather(mismatch_rows, key = "Variable", value = "Check_Value", all_of(check_columns))
mismatch_rows_long <- mismatch_rows_long[mismatch_rows_long$Check_Value =='Mismatch',] 

variables <- unique(mismatch_rows_long$Variable)
for (variable in variables) {
  variable_name <- gsub("^Check_", "", variable)
  xx <- mismatch_rows_long[mismatch_rows_long$Variable == variable, ]
  mismatch_rows_long[mismatch_rows_long$Variable == variable,'diff'] <-
    xx[, paste0(variable_name, ".y")] - xx[, paste0(variable_name, ".x")]
}

mismatch_rows_long_indextot <- mismatch_rows_long[mismatch_rows_long$Variable == 'Check_Index.Total',]

mismatch_rows_long_indextot_long <- tidyr::pivot_longer(mismatch_rows_long_indextot, 
                                                       cols = c("Index.Total.x", "Index.Total.y"), 
                                                       names_to = "Category", 
                                                       values_to = "Value")

plot <- ggplot(mismatch_rows_long_indextot_long, aes(x = Category, y = Value, fill = Category)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Mismatch Plot for summed up of reported values vs County Total values",
       fill = "Category") +
  facet_wrap(~County + Year, scales = "free_y") +
  theme(legend.position = "none") +
  ylim(0, max(mismatch_rows_long_indextot$Index.Total.x, mismatch_rows_long_indextot$Index.Total.y) + 500)

print(plot)

mismatch_rows_long_indextot <- subset(mismatch_rows_long_indextot, County=='Albany')

# ggplot(mismatch_rows_long_indextot, aes(x = interaction(Year, County))) +
#   geom_bar(aes(y = Index.Total.y, fill = "County Total values"), stat = "identity") +
#   geom_bar(aes(y = Index.Total.x, fill = "Sum of Reported Values by Agencies"), stat = "identity") +
#   labs(title = "Difference between County total values and summed up provided values for Albany county", 
#        x = "Year", 
#        y = "Total Number of Crimes") +
#   scale_fill_manual(values = c("County Total values" = "red", 
#                                "Sum of Reported Values by Agencies" = "blue")) +
#   theme_minimal()+
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

ggplot(mismatch_rows_long_indextot, 
       aes(x = interaction(Year, County))) +
  geom_bar(aes(y = Index.Total.y, 
               fill = "County Total values"), 
           stat = "identity", 
           position = "dodge") +
  geom_bar(aes(y = Index.Total.x, 
               fill = "Sum of Reported Values by Agencies"), 
           stat = "identity", 
           position = "dodge") +
  labs(title = "Difference between County total values and summed up provided values for Albany county", 
       x = "Year + County", 
       y = " Number of Crimes") +
  scale_fill_manual(values = c("County Total values" = "red", 
                               "Sum of Reported Values by Agencies" = "blue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

#subset_data <- mismatch_rows_long[mismatch_rows_long$Variable == 'Check_Index.Total', ]
#subset_data$diff <- subset_data$Index.Total.x - subset_data$Index.Total.y

```




```{r}
print(sum(crime_ny[(crime_ny$County=='Nassau') & (crime_ny$Year == 1990) & (crime_ny$Agency != 'County Total'),]$Index.Total))
print(crime_ny[(crime_ny$County=='Nassau') & (crime_ny$Year == 1990) & (crime_ny$Agency == 'County Total'),]$Index.Total)

crime_ny[(crime_ny$County=='Albany') & (crime_ny$Year == 1990) & (crime_ny$Agency != 'County Total'),]$Index.Total
crime_ny[(crime_ny$County=='Albany') & (crime_ny$Year == 1991) & (crime_ny$Agency != 'County Total'),]$Index.Total

crime_ny[(crime_ny$County=='Albany') & (crime_ny$Year == 1990) & (crime_ny$Agency == 'County Total'),]
crime_ny[(crime_ny$County=='Albany') & (crime_ny$Year == 1991) & (crime_ny$Agency == 'County Total'),]

nrow(crime_ny[(crime_ny$County=='Nassau') & (crime_ny$Year == 1991) & (crime_ny$Agency != 'County Total'),])==nrow(crime_ny[(crime_ny$County=='Nassau') & (crime_ny$Year == 1990) & (crime_ny$Agency != 'County Total'),])

for (year in 1990:2022) {
 cat(year,' and ',nrow(crime_ny[(crime_ny$County=='Albany') & (crime_ny$Year == year) & (crime_ny$Agency != 'County Total'),]),'\n')
}

alb_1997 <- unique(crime_ny[(crime_ny$County=='Albany') & (crime_ny$Year == 1997) & (crime_ny$Agency != 'County Total'),]$Agency)
alb_1998 <- unique(crime_ny[(crime_ny$County=='Albany') & (crime_ny$Year == 1998) & (crime_ny$Agency != 'County Total'),]$Agency)

for (ag in alb_1997) {
  if (!(ag %in% alb_1998)){
    print(ag)
  }
}

crime_ny[(crime_ny$Agency=="Albany County (RR-D and H)"),]


```


```{r}
# Check that other entries sum to Index.Total:

# First check Property Total + Violent Total == Index Total
check_result <- crime_ny %>%
  group_by(Year, County, Agency) %>%
  summarise(Sum_Other_Entries = Property.Total + Violent.Total) %>%
  left_join(crime_ny, by = c("Year", "County", "Agency")) %>%
  mutate(Check = ifelse(Sum_Other_Entries == Index.Total, "Pass", "Fail"))
cat('Number of rows with mismatch of Property Total + Violent Total == Index Total = ',nrow(check_result[check_result$Check!='Pass',]),'\n')

# First check Murder + Rape + Robbery + Aggravated Assault == Violent Total
check_result <- crime_ny %>%
  group_by(Year, County, Agency) %>%
  summarise(Sum_Other_Entries = Murder + Rape + Robbery + Aggravated.Assault) %>%
  left_join(crime_ny, by = c("Year", "County", "Agency")) %>%
  mutate(Check = ifelse(Sum_Other_Entries == Violent.Total, "Pass", "Fail"))
print(nrow(check_result[check_result$Check!='Pass',]))
cat('Number of rows with mismatch of Murder + Rape + Robbery + Aggravated Assault == Violent Total = ',nrow(check_result[check_result$Check!='Pass',]),'\n')

# First check Burglary + Larceny + Motor.Vehicle.Theft == Property Total
check_result <- crime_ny %>%
  group_by(Year, County, Agency) %>%
  summarise(Sum_Other_Entries = Burglary + Larceny + Motor.Vehicle.Theft) %>%
  left_join(crime_ny, by = c("Year", "County", "Agency")) %>%
  mutate(Check = ifelse(Sum_Other_Entries == Property.Total, "Pass", "Fail"))
cat('Number of rows with mismatch of Burglary + Larceny + Motor.Vehicle.Theft == Property Total Total = ',nrow(check_result[check_result$Check!='Pass',]),'\n')

```

## Some preliminary graphs

```{r}

# kinda useless graphics

ggplot(crime_ny, aes(x = County, y = Index.Total, fill = Index.Total)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Year, scales = "free") +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Index.Total",
       x = "County",
       y = "Index.Total")
```

### Total crime yearly breakdown in NY stat

```{r}

# total crime per county per year Interesting upswing in 2022 which we should investigate.

crime_ny$Year <- as.numeric(crime_ny$Year)

crime_ny_filtered <- crime_ny %>%
  mutate(Year_Period = cut(Year, 
                           breaks = seq(1985, max(Year) + 5, 
                           by = 5), 
                           labels = FALSE),
         Year_Period_Label = paste0(floor((Year - 1985) / 5) * 5 + 1985,
                                    "-",
                                    floor((Year - 1985) / 5) * 5 + 1989)
         )

crime_ny_filtered_county_tot <- crime_ny_filtered[crime_ny_filtered$Agency=='County Total',]

tot_yearly <- crime_ny_filtered_county_tot %>%
  group_by(Year) %>%
  summarise(
    Total_Index = sum(Index.Total)
  ) %>%
  ungroup()

ggplot(tot_yearly, aes(x = Year, 
                      y = Total_Index, 
                      fill = Total_Index)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Index.Total by County (Faceted by 5-Year Intervals)",
       x = "County",
       y = "Index.Total")

```

### 5 year total crime by state (top 5 states in each period)

```{r}

# faceted 5-year period / county stats

combined_data <- crime_ny_filtered_county_tot %>%
  group_by(Year_Period_Label, County) %>%
  summarise(
    Total_Index = sum(Index.Total),
  ) %>%
  ungroup()

top_5_counts <- combined_data %>%
  group_by(Year_Period_Label) %>%
  top_n(5, wt = Total_Index) %>%
  ungroup()

top_5_counts$County <- fct_reorder(top_5_counts$County, 
                                   top_5_counts$Total_Index, 
                                   .desc = TRUE)

top_5_counts <- top_5_counts %>%
  group_by(Year_Period_Label) %>%
  mutate(Rank = -rank(Total_Index))

top_5_counts$County <- fct_reorder(top_5_counts$County, top_5_counts$Rank)


ggplot(top_5_counts, aes(x = County, 
                         y = Total_Index, 
                         fill = Total_Index)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Year_Period_Label, 
             scales = "free", #"free_x",
             nrow = 5) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Index.Total by County",
       x = "County",
       y = "Index.Total")
```

### Draw heat map maps

```{r}

county_fips <- data.frame(
  FIPS = c(
    "001", "005", "007", "009", "011", "013", "015", "017", "019", "021",
    "023", "025", "027", "029", "031", "033", "035", "037", "039", "041",
    "043", "045", "047", "049", "051", "053", "055", "057", "059", "061",
    "063", "065", "067", "069", "071", "073", "075", "077", "079", "081",
    "083", "085", "087", "089", "091", "093", "095", "097", "099", "101",
    "103", "105", "107", "109", "111", "113", "115", "117", "119", "121", "123"
  ),
  Name = c(
    "Albany", "Bronx", "Broome", "Cattaraugus", "Cayuga", "Chautauqua",
    "Chemung", "Chenango", "Clinton", "Columbia", "Cortland", "Delaware",
    "Dutchess", "Erie", "Essex", "Franklin", "Fulton", "Genesee", "Greene",
    "Hamilton", "Herkimer", "Jefferson", "Kings", "Lewis", "Livingston",
    "Madison", "Monroe", "Montgomery", "Nassau", "New York", "Niagara",
    "Oneida", "Onondaga", "Ontario", "Orange", "Orleans", "Oswego",
    "Otsego", "Putnam", "Queens", "Rensselaer", "Richmond", "Rockland",
    "St. Lawrence", "Saratoga", "Schenectady", "Schoharie", "Schuyler",
    "Seneca", "Steuben", "Suffolk", "Sullivan", "Tioga", "Tompkins",
    "Ulster", "Warren", "Washington", "Wayne", "Westchester", "Wyoming", "Yates"
  )
)
crime_ny$County <- as.character(crime_ny$County)
crime_ny_with_fips <- merge(crime_ny, county_fips, by.x = "County", by.y = "Name", all.x = TRUE)
crime_ny_with_fips <- crime_ny_with_fips[crime_ny_with_fips$Agency == 'County Total', ]

ny_state_boundaries <- st_read("../../tl_2016_36_cousub/tl_2016_36_cousub.shp")
ny_state_boundaries <- st_transform(ny_state_boundaries, crs = 4326)
#plot(ny_state_boundaries)t

ny_state_boundaries$COUNTYFP <- as.character(ny_state_boundaries$COUNTYFP)
crime_ny_with_fips$FIPS <- as.character(crime_ny_with_fips$FIPS)

crime_ny_merged <- merge(ny_state_boundaries, crime_ny_with_fips, by.x = "COUNTYFP", by.y = "FIPS")
crime_ny_sf <- st_as_sf(crime_ny_merged, coords = c("Longitude", "Latitude"), crs = 4326)

xx <- subset(crime_ny_sf, Year == 2020)

ggplot(xx, aes(fill = Index.Total)) +
  geom_sf(color = "black") +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = paste("Index.Total by County (Year:", 2020, ")"),
       fill = "Index.Total")

```

```{r}


filtered_data <- crime_ny_sf # subset(crime_ny_sf, (Year < 2015) & (Year > 2010))

plots <- lapply(unique(filtered_data$Year), function(year) {
ggplot(subset(filtered_data, Year == year), aes(fill = Index.Total)) +
  geom_sf(color = "black") +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = paste("Index.Total by County (Year:", year, ")"),
       fill = "Index.Total")
})

#plots

```

### Animate image --\> create gif (takes a while to evaluate)

```{r}
library(ggplot2)
library(gganimate)
library(viridis)
library(magick)

ny_state_boundaries <- st_read("../../tl_2016_36_cousub/tl_2016_36_cousub.shp")
ny_state_boundaries <- st_transform(ny_state_boundaries, crs = 4326)
#plot(ny_state_boundaries)t

ny_state_boundaries$COUNTYFP <- as.character(ny_state_boundaries$COUNTYFP)
crime_ny_with_fips$FIPS <- as.character(crime_ny_with_fips$FIPS)

crime_ny_merged <- merge(ny_state_boundaries, crime_ny_with_fips_and_pop, by.x = "COUNTYFP", by.y = "FIPS")
crime_ny_merged$crimeperpop <- crime_ny_merged$Index.Total/crime_ny_merged$Population
crime_ny_sf <- st_as_sf(crime_ny_merged, coords = c("Longitude", "Latitude"), crs = 4326)

#crime_ny_with_fips_and_pop

desired_crs <- 4326
filtered_data <- crime_ny_sf
dir.create("animation_frames", showWarnings = FALSE)

legend_breaks <- seq(min(filtered_data$crimeperpop), max(filtered_data$crimeperpop), length.out = 5)
legend_labels <- scales::number_format(scale = 100, suffix = "%")(legend_breaks)

scale_limits <- c(0, max(filtered_data$crimeperpop) * .5)

template_plot <- ggplot(filtered_data, aes(fill = crimeperpop)) +
  geom_sf(color = "black") +
  scale_fill_viridis_c(breaks = legend_breaks, 
                       labels = legend_labels, 
                       option = "viridis_div",
                       limits = scale_limits) +  # Set limits for the color scale
  theme_minimal() +
  labs(title = "Crime/pop by County", fill = "Crime/Pop")

ggsave("animation_frames/template_plot.png", 
       template_plot, 
       width = 6, 
       height = 6, 
       dpi = 100)

for (year in unique(filtered_data$Year)) {
  plot_data <- subset(filtered_data, Year == year)
  plot <- template_plot + 
    geom_sf(data = plot_data) +
    labs(title = paste("Crime/Pop by County (Year:", year, ")"))
  
  ggsave(filename = paste0("animation_frames/frame_", year, ".png"), plot, width = 6, height = 6, dpi = 100)
}

png_files <- list.files("animation_frames", pattern = "frame_\\d+\\.png", full.names = TRUE)
frames <- image_read(png_files)

animation_file <- "animated_plot.gif"
fps <- 2
width <- 800
height <- 800

animation <- image_animate(frames, fps = fps)
image_write(animation, path = animation_file)

#image_write(frames, path = animation_file, format = "gif", fps = 1)
#unlink("animation_frames", recursive = TRUE)

# desired_crs <- 4326
# filtered_data <- crime_ny_sf
# dir.create("animation_frames", showWarnings = FALSE)
# 
# for (year in unique(filtered_data$Year)) {
#   plot_data <- subset(filtered_data, Year == year)
#   plot <- ggplot() +
#     geom_sf(data = plot_data, aes(fill = Index.Total), color = "black") +
#     scale_fill_viridis_c() +
#     theme_minimal() +
#     labs(title = paste("Index.Total by County (Year:", year, ")"), fill = "Index.Total")
#   
#   ggsave(filename = paste0("animation_frames/frame_", year, ".png"), plot, width = 6, height = 6, dpi = 100)
# }
# 
# png_files <- list.files("animation_frames", pattern = "frame_\\d+\\.png", full.names = TRUE)
# frames <- image_read(png_files)
# 
# animation_file <- "animated_plot.gif"
# fps <- 2
# width <- 800
# height <- 800
# 
# animation <- image_animate(frames, fps = fps)
# image_write(animation, path = animation_file)

#image_write(frames, path = animation_file, format = "gif", fps = 1)
#unlink("animation_frames", recursive = TRUE)
```

```{r}

# crime_ny_sf$Year <- as.numeric(crime_ny_sf$Year)
# 
# animated_plot <- ggplot(filtered_data, aes(fill = Index.Total)) +
#   geom_sf(color = "black") +
#   scale_fill_viridis_c() +
#   theme_minimal() +
#   transition_states(Year, 
#                     transition_length = 1, 
#                     state_length = 1, 
#                     wrap = TRUE) +
#   enter_fade() +
#   exit_fade() +
#   labs(title = "Index.Total by County (Year: {frame_time})", fill = "Index.Total")
# 
# 
# anim_save("animated_plot.gif", animated_plot, renderer = gifski_renderer())

```

```{r}
#some checks:
# st_crs(ny_state_boundaries)==st_crs(crime_ny_sf)
# print(sum(st_is_empty(ny_state_boundaries)))
# print(sum(st_is_empty(crime_ny_sf)))
# print(sum(!st_is_valid(ny_state_boundaries)))
# print(sum(!st_is_valid(crime_ny_sf)))
```

------------------------------------------------------------------------

# NYC shooting incident data

## Some data checks

```{r}
crime_nyc <- read.csv('../data/NYPD_Shooting_Incident_Data__Historic__20231124.csv')

crime_nyc$OCCUR_DATE <- as.Date(crime_nyc$OCCUR_DATE, format = '%m/%d/%Y')

na_cols <- colSums(is.na(crime_nyc))
cat("Columns with NA values:", names(na_cols[na_cols>0]), "\n")

# drop juristiction code alltogether

crime_nyc <- subset(crime_nyc, select = -JURISDICTION_CODE)
na_cols <- colSums(is.na(crime_nyc))
na_rows <- crime_nyc[is.na(crime_nyc[, names(na_cols[na_cols>0])]), ]

print(head(na_rows,30))

# mostly empty rows -- two incidences have missing jurisdiction_code, but we don't really care about that.

#cat("Rows with NA values:", nrow(subset(na_rows, !(Agency == "County Total"))), "\n")
#cat("Newest with NA values:",max((subset(na_rows, !(Agency == "County Total")))$Year), "\n")

head(crime_nyc,5)
```

```{r}
missing_percentage <- sapply(crime_nyc, 
                             function(x) sum(is.na(x) | x == "") / length(x) * 100)
result <- data.frame(Column = names(missing_percentage), 
                     MissingPercentage = missing_percentage)
result <- result[order(-result$MissingPercentage), ]
print(result)
```

## preliminary graphs

## different maps:

### Leaflet plot:

```{r}

library(leaflet)
library(sf)

crime_nyc_nona <- crime_nyc[complete.cases(crime_nyc$Longitude, crime_nyc$Latitude), ]
nyc_map <- leaflet() %>%
  setView(lng = -73.94, lat = 40.70, zoom = 10) 

crime_nyc_sf <- st_as_sf(crime_nyc_nona, 
                         coords = c("Longitude", "Latitude"), 
                         crs = 4326)

crime_nyc_sf <- crime_nyc_sf

nyc_map <- nyc_map %>%
  addProviderTiles("CartoDB.Positron") %>%
  addCircleMarkers(
    data = crime_nyc_sf, 
    color = "blue", 
    radius = 0, 
    opacity = 0.3,
    fillOpacity = 1,
    stroke = FALSE,
    popup = ~paste("Date: ", crime_nyc_sf$OCCUR_DATE, "<br>", "Time: ", crime_nyc_sf$OCCUR_TIME)
  )

# Display the map
nyc_map

```

### Various heatmaps (different technique):

```{r}

library(ggplot2)
library(sp)
library(sf)

crime_nyc <- read.csv('../data/NYPD_Shooting_Incident_Data__Historic__20231124.csv')
crime_nyc$OCCUR_DATE <- as.Date(crime_nyc$OCCUR_DATE, format = '%m/%d/%Y')

na_cols <- colSums(is.na(crime_nyc))
cat("Columns with NA values:", names(na_cols[na_cols>0]), "\n")

crime_nyc <- subset(crime_nyc, select = -JURISDICTION_CODE)
na_cols <- colSums(is.na(crime_nyc))
na_rows <- crime_nyc[is.na(crime_nyc[, names(na_cols[na_cols>0])]), ]

print(head(na_rows,30))

# mostly empty rows -- two incidences have missing jurisdiction_code, but we don't really care about that.

#cat("Rows with NA values:", nrow(subset(na_rows, !(Agency == "County Total"))), "\n")
#cat("Newest with NA values:",max((subset(na_rows, !(Agency == "County Total")))$Year), "\n")


dsn <- "nyu_2451_34490.shp"
nyc_boundaries <- st_read('../../nyc_election_districts/nyc_election_districts.shp')

#plot(nyc_boundaries)

crime_nyc_sf <- st_as_sf(crime_nyc_nona, coords = c("Longitude", "Latitude"), crs = 4326)

ggplot() +
  geom_sf(data = nyc_boundaries, fill = "white", color = "black") +
  geom_point(data = crime_nyc_nona, aes(x = Longitude, y = Latitude), 
             color = "yellow", 
             size = 0.01, 
             alpha = 0.08) +
  #geom_density_2d(data = crime_nyc_nona, aes(x = Longitude, y = Latitude)) +
  geom_density_2d_filled(data = crime_nyc_nona, aes(x = Longitude, y = Latitude),alpha = 0.85) +
  theme_minimal()

```

```{r}

#last 10 yrs

crime_nyc_nona$OCCUR_DATE <- as.Date(crime_nyc_nona$OCCUR_DATE, format = '%m/%d/%Y')
crime_nyc_nona_last10 <- subset(crime_nyc_nona, OCCUR_DATE >= as.Date('2022-01-01'))

crime_nyc_sf <- st_as_sf(crime_nyc_nona_last10, coords = c("Longitude", "Latitude"), crs = 4326)

ggplot() +
  geom_sf(data = nyc_boundaries, fill = "white", color = "black") +
  #geom_density_2d(data = crime_nyc_nona, aes(x = Longitude, y = Latitude)) +
  geom_density_2d_filled(data = crime_nyc_nona_last10, aes(x = Longitude, y = Latitude),alpha = 0.85) +
  theme_minimal()
```

```{r}
ggplot() +
  geom_sf(data = nyc_boundaries, fill = "white", color = "black") +
  geom_point(data = crime_nyc_nona_last10, aes(x = Longitude, y = Latitude), 
             color = "red", 
             size = .5, 
             alpha = 0.4)
```
