
<!-- 

II. D. Data
The data can be pulled from multiple sources; it does not need to be a single dataset. Be sure to get data from the original source. For example, if you wish to work with data collected and distributed by the  Centers for Disease Control, that is where you should go to access the data, not a third party that has posted the data. Do not use datasets that have been processed and cannot be traced to the source. For this reason, you may not use datasets from Kaggle. We encourage a diverse set of projects; for this reason you cannot use datasets that we've seen over and over again: Citibike, Yelp, NYC restaurant ratings, Airbnb, and others. If there's an issue with your data, we will let you know after you turn in Data check -- optional  or PSet 4 -- Final project checkpoint. In addition you can ask during office hours before then.

----------------------------------------------------------------------

  2.1 Technical description

Identify one or more data sources (see II. D. above) that you propose to draw on for the project. For each, describe how the data are collected and by whom. Describe the format of the data, the frequency of updates, dimensions, and any other relevant information. Note any issues / problems with the data, either known or that you discover. Explain how you plan to import the data. Carefully document your sources with links to the precise data sources that you used. If that is not possible (for example if your data is not available online, then explain that clearly.)

(suggested: 1 page)

  2.2 Research plan

Explain in detail how your data will enable you to answer your research questions from the introduction. 

(suggested: 1 page)

 2.3 Missing value analysis

Describe any patterns you discover in missing values. If no values are missing, graphs should still be included showing that.

(suggested: 2 graphs plus commentary)

-->

# Data

## Technical Description

The dataset on [Index Crimes by County and Agency in New York](https://data.ny.gov/Public-Safety/Index-Crimes-by-County-and-Agency-Beginning-1990/ca8h-8gjq), is collected by the New York State Division of Criminal Justice Services, which obtains crime reports from more than 500 New York State police and sheriffs' departments. The data is formatted in a tabular structure, with columns specifying details like county, agency, year, region, and various crime categories. So, the dataset includes geographic information (county and agency), temporal information (year), and crime categories, and contains roughly 22 thousand rows and 15 columns.

The reporting system gathers data on seven Index offenses, which are commonly utilized to assess the overall volume of crime. These encompass violent crimes such as murder/non-negligent manslaughter, forcible rape, robbery, and aggravated assault, as well as property crimes like burglary, larceny, and motor vehicle theft.

The data has been collected since January 1990. The frequency of updates is on an annual basis, but the individual law enforcement agencies monthly report their total crimes to the New York State Division of Criminal Justice Services.
Possible issues with the data could include inconsistencies in reporting methods among different agencies, variations in the definition of crimes, or potential underreporting. We should be cautious about these nuances when interpreting trends or making comparisons across regions and time periods. Indeed, police agencies might encounter challenges in reporting, which could hinder accurate or comprehensive data submission. The figures reflect solely the crimes reported to the police and do not encompass the total number of crimes that took place.

We carefully check that the data is consistent, i.e. that the various accumulations performed within the dataset are correct. We find some inconsistencies in the County Total accumulation in Albany from 1990-1997 and Nassau County in 1990. We discuss this further in the missing values analysis.

Since comparing crime numbers among the different counties only really makes sense if we normalize by population, we supplement our main dataset, [Index Crimes by County and Agency in New York](https://data.ny.gov/Public-Safety/Index-Crimes-by-County-and-Agency-Beginning-1990/ca8h-8gjq), with an official population dataset of the counties in NY state. The [population dataset](https://data.ny.gov/Government-Finance/Annual-Population-Estimates-for-New-York-State-and/krt9-ym2k), is a dataset curated by the New York State Department of Labor and started in 2014. It has been updated yearly since then, and contains the (estimated) population of each County in NY state by year since 1970. 

The dataset contains the [FIPS](https://en.wikipedia.org/wiki/Federal_Information_Processing_Standards) ("Federal Information Processing Standards") code of each NY state county together with the name of the county ("Geography"), the year and its population estimate. Finally, it includes a column, "Program Type", detailing the way the population number was estimated. The estimations are interpolations or extrapolations of the 10 year census data of NY State. Before 2010, the estimations are based on interpolations, since for each year in between the actual census data, both previous and next census numbers were known. However, after 2010, the data was kept as point-in-time, and therefore the estimations were based solely on extrapolating the previous census numbers. Notice that it is important to keep the data point-in-time as changing the numbers in hindsight would lead to issues with models that use the extrapolated data numbers.

The more detailed methodology of the "intercensal estimation" are detailed [here](https://www.epa.gov/sites/default/files/2015-05/documents/aceappendixb_censusintercensalandpostcensaldata.pdf) and [here](https://www2.census.gov/programs-surveys/popest/technical-documentation/methodology/intercensal/2000-2010-intercensal-estimates-methodology.pdf). Essentially, for intercensal estimates, they combine the data from the two censuses with death, birth, migration and immigration data to provide an estimate, while for postcensal estimates, they use the previous census data together with the same external sources to provide an estimate.

We shall investigate the efficacies of these estimates briefly below and notice that the extrapolation is comparatively worse than the intercensal estmated data.

## Research plan

To answer research questions related to crime trends and influencing factors, the first dataset offers a rich source of information. Regarding the temporal analysis, we can examine crime rates over multiple years and then can identify trends and patterns. This enables the assessment of whether crime has increased, decreased, or remained stable over time, providing insights into the effectiveness of law enforcement strategies, changes in societal conditions or from the perspective of pre/during/post covid.
Moreover, a geographic analysis is also possible. Indeed, the dataset includes information on crimes reported by different counties and agencies. This allows us to compare crime rates across regions, potentially revealing spatial patterns. For example, we can compare the evolution of crime between NYC and NY state. Understanding variations in crime rates helps tailor interventions to specific areas. 
The dataset is categorized into specific crime types. So, we can analyze trends in each category individually, helping to identify areas of concern and inform targeted interventions. 
Furthermore, additional demographic information, if available, could be integrated into the analysis. This might involve exploring correlations between crime rates and socioeconomic factors, population density, or demographic shifts over time. 
We can also assess the impact of law enforcement strategies by analyzing changes in crime rates following the implementation of specific policies or practices.
Since the dataset spans several years, we can conduct longitudinal studies, observing changes and trends over extended periods. This can provide a more nuanced understanding of the factors influencing crime dynamics. 
By comparing data across different agencies or counties, researchers can identify variations in reporting practices, law enforcement effectiveness, or community characteristics that may influence crime rates. 
However, it's crucial to acknowledge potential limitations such as data inconsistencies, reporting biases, and missing information, which may affect the robustness of conclusions drawn from the analysis.





## Missing value analysis and other inconsistencies

We shall now investigate our two datasets for missing values, inconsistencies and discuss if our findings will impact our research and results.

### NY State crime dataset

We start by investigating the NY State crime dataset.

```{r, echo=FALSE,results='hide', message=FALSE, warning=FALSE}
library(sf)
library(dplyr)
library(ggplot2)
library(patchwork)
library(gifski)
library(transformr)
library(dplyr)
library(ggplot2)
library(forcats)
library(gganimate)

# load and clean/fix data for later use:

crime_ny <- read.csv("data/Index_Crimes_by_County_and_Agency__Beginning_1990.csv")

# not planing on using this:
# crime_nyc <- read.csv('data/NYPD_Shooting_Incident_Data__Historic__20231124.csv')

pop_ny <- read.csv('data/Annual_Population_Estimates_for_New_York_State_and_Counties__Beginning_1970.csv')

# standardize FIPS to work with below conventions:
pop_ny$FIPS.Code <- substr(pop_ny$FIPS.Code, 3, nchar(pop_ny$FIPS.Code))

# map fips to County name:
county_fips <- data.frame(
  FIPS = c(
    "001", "003", "005", "007", "009", "011", "013", "015", "017", "019", "021",
    "023", "025", "027", "029", "031", "033", "035", "037", "039", "041",
    "043", "045", "047", "049", "051", "053", "055", "057", "059", "061",
    "063", "065", "067", "069", "071", "073", "075", "077", "079", "081",
    "083", "085", "087", "089", "089", "091", "093", "095", "097", "099", "101",
    "103", "105", "107", "109", "111", "113", "115", "117", "119", "121", "123"
  ),
  Name = c(
    "Albany", "Allegany", "Bronx", "Broome", "Cattaraugus", "Cayuga", "Chautauqua",
    "Chemung", "Chenango", "Clinton", "Columbia", "Cortland", "Delaware",
    "Dutchess", "Erie", "Essex", "Franklin", "Fulton", "Genesee", "Greene",
    "Hamilton", "Herkimer", "Jefferson", "Kings", "Lewis", "Livingston",
    "Madison", "Monroe", "Montgomery", "Nassau", "New York", "Niagara",
    "Oneida", "Onondaga", "Ontario", "Orange", "Orleans", "Oswego",
    "Otsego", "Putnam", "Queens", "Rensselaer", "Richmond", "Rockland",
    "St. Lawrence", "St Lawrence", "Saratoga", "Schenectady", "Schoharie", "Schuyler",
    "Seneca", "Steuben", "Suffolk", "Sullivan", "Tioga", "Tompkins",
    "Ulster", "Warren", "Washington", "Wayne", "Westchester", "Wyoming", "Yates"
  )
)

# add fips to crime_ny --> will allow us to identify with fips in .shp file and merge with pop_ny file.
crime_ny$County <- as.character(crime_ny$County)
crime_ny_with_fips <- merge(crime_ny, county_fips, by.x = "County", by.y = "Name", all.x = TRUE)
crime_ny_with_fips <- crime_ny_with_fips[crime_ny_with_fips$Agency == 'County Total', ]

# merge population and NY-state crime datasets:
crime_ny_with_fips_and_pop <- merge(crime_ny_with_fips, 
                                    pop_ny, by.x = c("FIPS", "Year"), 
                                    by.y = c("FIPS.Code", "Year"), 
                                    all.x = TRUE)
crime_ny_with_fips_and_pop <- crime_ny_with_fips_and_pop[!is.na(crime_ny_with_fips_and_pop$Index.Total), ]

# drop geography and program type from merged dataframe (geograph is simply the name of the county)
# and program type is the way it was estimated.
crime_ny_with_fips_and_pop <- crime_ny_with_fips_and_pop[, !(names(crime_ny_with_fips_and_pop) %in% c("Geography", "Program.Type"))]

# check NA columns for crime_ny:
na_cols <- colSums(is.na(crime_ny_with_fips_and_pop))
cat("Columns with NA values:", names(na_cols[na_cols>0]), "\n")
# find NA values only in Months Reported column.
```

We find that the data is really well-sourced and only contains NA values for the "Months Reported" column. In particular, upon further investigation, it turns out that a lot of the NA values appear in the accumulated County Total rows, because as soon as an Agency does not indicate the number of months they reported this will lead to an NA value in the accumulation.

Otherwise, below we have summarized the percentage of NA values in the Months Collected column per county.

```{r}
na_cols <- colSums(is.na(crime_ny))
na_rows <- crime_ny[is.na(crime_ny[, names(na_cols[na_cols>0])]), ]

na_rows_cnt <- na_rows %>% group_by(County) %>% summarise(Count = n())
crime_ny_cnt <- crime_ny %>% group_by(County) %>% summarise(Count = n())
na_rows_cnt$pct <- na_rows_cnt$Count/crime_ny_cnt$Count

na_rows_cnt <- na_rows_cnt[order(na_rows_cnt$pct, decreasing = TRUE), ]

ggplot(na_rows_cnt, aes(x = reorder(County, -pct), y = pct)) +
  geom_bar(stat = "identity") +
  labs(title = "Percentage of Missing \"Months Collected\" Values by County",
       x = "County", y = "Percentage") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

We notice that the five counties with the most missing value percentages in the Months Collected column are the ones, where we do not have Agency-level data, only County total data. We believe these numbers were aggregated from different datasets, which may have varying degree of accuracy with respect to their Months collected information.

While we are not specifically interested in the Months collected column, we need to keep in mind that if for strictly less than 12 months per year data was collected, there might be a skew in the data. We therefore have a quick look at the strictly less than 12 months reported within the non-NA reported values.


```{r}

less_12 <- crime_ny[(crime_ny$Months.Reported < 12) & !is.na(crime_ny$Months.Reported), ]
less_12_cnt <- less_12 %>% group_by(County) %>% summarise(Count = n())

no_na_crime_ny_cnt <- crime_ny[!is.na(crime_ny$Months.Reported), ] %>% group_by(County) %>% summarise(Count = n())
merged_data <- merge(less_12_cnt, no_na_crime_ny_cnt, by = "County", all.x = TRUE)
merged_data$pct <- merged_data$Count.x/merged_data$Count.y

merged_data <- merged_data[order(merged_data$pct, decreasing = TRUE), ]

ggplot(merged_data, aes(x = reorder(County, -pct), y = pct)) +
  geom_bar(stat = "identity") +
  labs(title = "Percentage of <12 vs =12\"Months Collected\" by County",
       x = "County", 
       y = "Percentage of <12 vs =12 Months Collected") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

We see that the data with less than 12 months collected compared against the data with exactly 12 months collected is negligible. Based on this, we believe that the missing NA values are mostly due to data collection negligency rather than actually varying number of months collected.

Apart from the missing values, we wanted to check that the data aggregation was consistent. We first checked that the columns sum to their respective total columns for each row, and find that there is no inconsistency there:

```{r, warning = FALSE}
# Check that other entries sum to Index.Total:

# First check Property Total + Violent Total == Index Total
check_result <- crime_ny %>%
  group_by(Year, County, Agency) %>%
  summarise(Sum_Other_Entries = Property.Total + Violent.Total, .groups = "drop") %>%
  left_join(crime_ny, by = c("Year", "County", "Agency")) %>%
  mutate(Check = ifelse(Sum_Other_Entries == Index.Total, "Pass", "Fail"))
cat(nrow(check_result[check_result$Check!='Pass',]), ' rows with mismatch of Property Total + Violent Total == Index Total.\n')

# First check Murder + Rape + Robbery + Aggravated Assault == Violent Total
check_result <- crime_ny %>%
  group_by(Year, County, Agency) %>%
  summarise(Sum_Other_Entries = Murder + Rape + Robbery + Aggravated.Assault, .groups = "drop") %>%
  left_join(crime_ny, by = c("Year", "County", "Agency")) %>%
  mutate(Check = ifelse(Sum_Other_Entries == Violent.Total, "Pass", "Fail"))
cat(nrow(check_result[check_result$Check!='Pass',]), ' rows with mismatch of Murder + Rape + Robbery + Aggravated Assault == Violent Total. \n')

# First check Burglary + Larceny + Motor.Vehicle.Theft == Property Total
check_result <- crime_ny %>%
  group_by(Year, County, Agency) %>%
  summarise(Sum_Other_Entries = Burglary + Larceny + Motor.Vehicle.Theft, .groups = "drop") %>%
  left_join(crime_ny, by = c("Year", "County", "Agency")) %>%
  mutate(Check = ifelse(Sum_Other_Entries == Property.Total, "Pass", "Fail"))
cat(nrow(check_result[check_result$Check!='Pass',]), ' rows with mismatch of Burglary + Larceny + Motor.Vehicle.Theft == Property Total Total. \n')

```

Finally, we would also like to check whether there is a mismatch between the County Total accumulation and the individual (by-agency) values. It turns out that for the cases in which we have by-agency data, there is a very minor mismatch between the summed data and the County total data in Nassau in 1990 and in Albany between 1990 and 1997. We conjecture that it might be due to data from a missing agency, but could not verify this.

```{r}
# Find mismatch in County Total vs per-police station:
county_total_entries <- crime_ny %>%
  filter(Agency == "County Total")

columns_to_sum <- c(
  "Index.Total", "Violent.Total", "Murder", "Rape",
  "Robbery", "Aggravated.Assault", "Property.Total",
  "Burglary", "Larceny", "Motor.Vehicle.Theft"
)

crime_ny_not_ct <- crime_ny[crime_ny$Agency!='County Total',] 

summed_data <- crime_ny_not_ct %>%
  group_by(Year, County) %>%
  summarise(across(all_of(columns_to_sum), sum), .groups = "drop")

comparison_result <- left_join(summed_data, 
                               county_total_entries, 
                               by = c("Year", "County"))

comparison_result <- comparison_result %>%
  mutate(
    Check_Index.Total = ifelse(Index.Total.x == Index.Total.y, "Match", "Mismatch"),
    Check_Violent.Total = ifelse(Violent.Total.x == Violent.Total.y, "Match", "Mismatch"),
    Check_Murder = ifelse(Murder.x == Murder.y, "Match", "Mismatch"),
    Check_Rape = ifelse(Rape.x == Rape.y, "Match", "Mismatch"),
    Check_Robbery = ifelse(Robbery.x == Robbery.y, "Match", "Mismatch"),
    Check_Aggravated.Assault = ifelse(Aggravated.Assault.x == Aggravated.Assault.y, "Match", "Mismatch"),
    Check_Property.Total = ifelse(Property.Total.x == Property.Total.y, "Match", "Mismatch"),
    Check_Burglary = ifelse(Burglary.x == Burglary.y, "Match", "Mismatch"),
    Check_Larceny = ifelse(Larceny.x == Larceny.y, "Match", "Mismatch"),
    Check_Motor.Vehicle.Theft = ifelse(Motor.Vehicle.Theft.x == Motor.Vehicle.Theft.y, "Match", "Mismatch")
  )

check_columns <- paste0("Check_", columns_to_sum)
mismatch_rows <- comparison_result %>%
  rowwise() %>%
  filter(any(c_across(all_of(check_columns)) == "Mismatch"))

mismatch_rows_long <- tidyr::gather(mismatch_rows, key = "Variable", value = "Check_Value", all_of(check_columns))
mismatch_rows_long <- mismatch_rows_long[mismatch_rows_long$Check_Value =='Mismatch',] 

variables <- unique(mismatch_rows_long$Variable)
for (variable in variables) {
  variable_name <- gsub("^Check_", "", variable)
  xx <- mismatch_rows_long[mismatch_rows_long$Variable == variable, ]
  mismatch_rows_long[mismatch_rows_long$Variable == variable,'diff'] <-
    xx[, paste0(variable_name, ".y")] - xx[, paste0(variable_name, ".x")]
}

mismatch_rows_long_indextot <- mismatch_rows_long[mismatch_rows_long$Variable == 'Check_Index.Total',]

mismatch_rows_long_indextot_long <- tidyr::pivot_longer(mismatch_rows_long_indextot, 
                                                       cols = c("Index.Total.x", "Index.Total.y"), 
                                                       names_to = "Category", 
                                                       values_to = "Value")

#mismatch_rows_long_indextot <- subset(mismatch_rows_long_indextot, County=='Albany')

ggplot(mismatch_rows_long_indextot, 
       aes(x = interaction(Year, County))) +
  geom_bar(aes(y = Index.Total.y, 
               fill = "County Total values"), 
           stat = "identity", 
           position = "dodge") +
  geom_bar(aes(y = Index.Total.x, 
               fill = "Sum of Reported Values by Agencies"), 
           stat = "identity", 
           position = "dodge") +
  labs(title = "Difference between County total values and summed up provided values for Albany county", 
       x = "Year + County", 
       y = " Number of Crimes") +
  scale_fill_manual(values = c("County Total values" = "red", 
                               "Sum of Reported Values by Agencies" = "blue")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```

In the plot above we compare for each inconsistency the discrepancy in the total crime number. We can observe that the inconsistency is very minor and will not affect our analysis below.


### NY State population dataset


```{r}
pop_ny <- read.csv('data/Annual_Population_Estimates_for_New_York_State_and_Counties__Beginning_1970.csv')
pop_ny$FIPS.Code <- substr(pop_ny$FIPS.Code, 3, nchar(pop_ny$FIPS.Code))

missing_percentage <- sapply(pop_ny, 
                             function(x) sum(is.na(x) | x == "") / length(x) * 100)
result <- data.frame(Column = names(missing_percentage), 
                     MissingPercentage = missing_percentage)
result <- result[order(-result$MissingPercentage), ]
print(result)
```

na_cols <- colSums(is.na(pop_ny))
cat("Columns with NA values:", names(na_cols[na_cols>0]), "\n")

```

```{r}

pop_ny_full_state <- pop_ny[pop_ny$Geography == 'New York State', ]

population_plot <- ggplot(pop_ny_full_state, 
                          aes(x = Year, 
                              y = Population, 
                              fill = Program.Type)) +
  geom_col(position = "dodge") +
  labs(title = "Population per Year", 
       x = "Year", 
       y = "Population") +
  scale_fill_manual(values = c("Census Base Population" = "blue", 
                              "Postcensal Population Estimate" = "red",
                              'Intercensal Population Estimate' = 'orange')) +
  theme_minimal()

plot(population_plot)

```


```{r}
library(tidyr)


census_years <- unique(pop_ny_full_state[pop_ny_full_state$Program.Type == 'Census Base Population', ]$Year)

pop_ny_full_state_cens_yrs <- subset(pop_ny_full_state, pop_ny_full_state$Year %in% census_years)

pop_ny_full_state_cens_yrs_wide <- pivot_wider(pop_ny_full_state_cens_yrs, 
                       names_from = Program.Type, 
                       values_from = Population)
colnames(pop_ny_full_state_cens_yrs_wide)

names(pop_ny_full_state_cens_yrs_wide)

ggplot(pop_ny_full_state_cens_yrs_wide, aes(x = Year)) +
  geom_point(aes(y = `Census Base Population`, color = "Census Base Population"), 
             position = position_dodge(width = 0.4)) +
  geom_point(aes(y = `Intercensal Population Estimate`, color = "Intercensal Estimate"), 
             position = position_dodge(width = 0.2)) +
  geom_point(aes(y = `Postcensal Population Estimate`, color = "Postcensal Estimate"), 
             position = position_dodge(width = -0.2)) +
  labs(title = "A", x = "B", y = "C") +
  scale_color_manual(values = c("Census Base Population" = "blue", 
                                "Intercensal Estimate" = "orange",
                                "Postcensal Estimate" = "red"),
                     name = "Legend Title") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```






