# Results


## Overall crime numbers in NY State and the impact of COVID-10


```{r, echo=FALSE,results='hide', message=FALSE, warning=FALSE}
library(viridis)
library(sf)
library(dplyr)
library(ggplot2)
library(patchwork)
library(gifski)
library(transformr)
library(dplyr)
library(ggplot2)
library(forcats)
library(gganimate)
library(tidyverse)
library(redav)

crime_ny <- read.csv("data/Index_Crimes_by_County_and_Agency__Beginning_1990.csv")
pop_ny <- read.csv('data/Annual_Population_Estimates_for_New_York_State_and_Counties__Beginning_1970.csv')
pop_ny$FIPS.Code <- substr(pop_ny$FIPS.Code, 3, nchar(pop_ny$FIPS.Code))

pop_ny <- pop_ny[!((pop_ny$Year %in% c(1970,1980,1990,2000,2010,2020)) & 
(pop_ny$Program.Type != 'Census Base Population')),]

county_fips <- data.frame(
  FIPS = c(
    "001", "003", "005", "007", "009", "011", "013", "015", "017", "019", "021",
    "023", "025", "027", "029", "031", "033", "035", "037", "039", "041",
    "043", "045", "047", "049", "051", "053", "055", "057", "059", "061",
    "063", "065", "067", "069", "071", "073", "075", "077", "079", "081",
    "083", "085", "087", "089", "089", "091", "093", "095", "097", "099", "101",
    "103", "105", "107", "109", "111", "113", "115", "117", "119", "121", "123"
  ),
  Name = c(
    "Albany", "Allegany", "Bronx", "Broome", "Cattaraugus", "Cayuga", "Chautauqua",
    "Chemung", "Chenango", "Clinton", "Columbia", "Cortland", "Delaware",
    "Dutchess", "Erie", "Essex", "Franklin", "Fulton", "Genesee", "Greene",
    "Hamilton", "Herkimer", "Jefferson", "Kings", "Lewis", "Livingston",
    "Madison", "Monroe", "Montgomery", "Nassau", "New York", "Niagara",
    "Oneida", "Onondaga", "Ontario", "Orange", "Orleans", "Oswego",
    "Otsego", "Putnam", "Queens", "Rensselaer", "Richmond", "Rockland",
    "St. Lawrence", "St Lawrence", "Saratoga", "Schenectady", "Schoharie", "Schuyler",
    "Seneca", "Steuben", "Suffolk", "Sullivan", "Tioga", "Tompkins",
    "Ulster", "Warren", "Washington", "Wayne", "Westchester", "Wyoming", "Yates"
  )
)
crime_ny$County <- as.character(crime_ny$County)
crime_ny_with_fips <- merge(crime_ny, county_fips, by.x = "County", by.y = "Name", all.x = TRUE)
crime_ny_with_fips <- crime_ny_with_fips[crime_ny_with_fips$Agency == 'County Total', ]

crime_ny_with_fips_and_pop <- merge(crime_ny_with_fips, 
                                    pop_ny, by.x = c("FIPS", "Year"), 
                                    by.y = c("FIPS.Code", "Year"), 
                                    all.x = TRUE)

crime_ny_with_fips_and_pop <- crime_ny_with_fips_and_pop[!is.na(crime_ny_with_fips_and_pop$Index.Total), ]
crime_ny_with_fips_and_pop <- crime_ny_with_fips_and_pop[, !(names(crime_ny_with_fips_and_pop) %in% c("Geography", "Program.Type"))]
```




```{r}
crime_ny$Year <- as.numeric(crime_ny$Year)
crime_ny_filtered <- crime_ny %>%
  mutate(Year_Period = cut(Year, 
                           breaks = seq(1985, max(Year) + 5, 
                           by = 5), 
                           labels = FALSE),
         Year_Period_Label = paste0(floor((Year - 1985) / 5) * 5 + 1985,
                                    "-",
                                    floor((Year - 1985) / 5) * 5 + 1989)
         )

crime_ny_filtered_county_tot <- crime_ny_filtered[crime_ny_filtered$Agency=='County Total',]

tot_yearly <- crime_ny_filtered_county_tot %>%
  group_by(Year) %>%
  summarise(
    Total_Index = sum(Index.Total)
  ) %>%
  ungroup()

ggplot(tot_yearly, aes(x = Year, 
                      y = Total_Index)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Total number of crimes in NY State by year",
       x = "Year",
       y = "Total number of recorded crimes")

```

We see an interesting uptick after 2021 potentially due to COVID-19. Worth investigating more!


```{r}

crime_ny_with_fips_and_pop$crime_pct <- crime_ny_with_fips_and_pop$Index.Total/crime_ny_with_fips_and_pop$Population

tot_yearly <- crime_ny_with_fips_and_pop %>%
  group_by(Year) %>%
  summarise(
    Total_crime = sum(Index.Total)
  ) %>%
  ungroup()

tot_yearly <- merge(tot_yearly, pop_ny[pop_ny$Geography == 'New York State', ], by.x = "Year", by.y = "Year", all.x = TRUE)

tot_yearly$pct <- tot_yearly$Total_crime/tot_yearly$Population

ggplot(tot_yearly, aes(x = Year, 
                      y = pct)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Crimes per population of crimes in NY State by year",
       x = "Year",
       y = "Recorded crimes per population")
```


### Increase in crimes per population per county


```{r}
impact_of_covid <- crime_ny_with_fips_and_pop
impact_of_covid$crime_pct <- impact_of_covid$Index.Total/impact_of_covid$Population

crime_increase <- impact_of_covid[impact_of_covid$Year > 2020, ] %>%
  group_by(County) %>%
  summarize(increase = last(crime_pct) - first(crime_pct)) %>%
  filter(!is.na(increase))

top_counties <- crime_increase %>% top_n(62, wt = increase)

selected_data <- impact_of_covid %>%
  filter(County %in% top_counties$County)

county_order <- names(sort(tapply(selected_data$crime_pct, selected_data$County, max), decreasing = TRUE))
selected_data$County <- factor(selected_data$County, levels = county_order)

my_palette <- viridis_pal()(length(unique(selected_data$County)))

ggplot(selected_data, aes(x = Year, y = crime_pct, color = County)) +
  geom_line() +
  scale_color_manual(values = my_palette) +
  labs(title = "Crime/population per Year for Counties with worst increase post-covid",
       x = "Year",
       y = "Crime/population") +
  theme_minimal()

```




```{r}

impact_of_covid <- crime_ny_with_fips_and_pop
impact_of_covid$crime_pct <- impact_of_covid$Index.Total/impact_of_covid$Population
impact_of_covid <- impact_of_covid[impact_of_covid$Year > 2015, ]

crime_increase <- impact_of_covid[impact_of_covid$Year > 2020, ] %>%
  group_by(County) %>%
  summarize(increase = last(crime_pct) - first(crime_pct)) %>%
  filter(!is.na(increase))

top_counties <- crime_increase %>% top_n(10, wt = increase)

selected_data <- impact_of_covid %>%
  filter(County %in% top_counties$County)

county_order <- names(sort(tapply(selected_data$crime_pct, selected_data$County, max), decreasing = TRUE))
selected_data$County <- factor(selected_data$County, levels = county_order)

my_palette <- viridis_pal()(length(unique(selected_data$County)))

ggplot(selected_data, aes(x = Year, y = crime_pct, color = County)) +
  geom_line() +
  scale_color_manual(values = my_palette) +
  labs(title = "Crime/population per Year for Counties with worst increase post-covid",
       x = "Year",
       y = "Crime/population") +
  theme_minimal()

```

```{r}
library(viridis)

impact_of_covid <- crime_ny_with_fips_and_pop
impact_of_covid$crime_pct <- impact_of_covid$Index.Total/impact_of_covid$Population
impact_of_covid <- impact_of_covid[impact_of_covid$Year > 2015, ]

bottom_counties <- crime_increase %>% top_n(10, wt = -increase)

selected_data <- impact_of_covid %>%
  filter((County %in% bottom_counties$County) | County %in% top_counties$County) 

county_order <- names(sort(tapply(selected_data$crime_pct, selected_data$County, max), decreasing = TRUE))
selected_data$County <- factor(selected_data$County, levels = county_order)

my_palette <- viridis_pal()(length(unique(selected_data$County)))

ggplot(selected_data, aes(x = Year, y = crime_pct, color = County)) +
  geom_line() +
  scale_color_manual(values = my_palette) +
  labs(title = "Crime/population per Year for Counties with least increase post-covid",
       x = "Year",
       y = "Crime/population") +
  theme_minimal()

```

Draw on map where each of these counties lie and see if there is a geographical connection.


```{r}
library(ggplot2)
library(gganimate)
library(viridis)
library(magick)

impact_of_covid <- crime_ny_with_fips_and_pop
impact_of_covid$crime_pct <- impact_of_covid$Index.Total/impact_of_covid$Population

crime_pct_increase <- impact_of_covid[impact_of_covid$Year >= 2020, ] %>%
  group_by(County) %>%
  summarize(increase = (last(crime_pct) - first(crime_pct))/first(crime_pct)) %>%
  filter(!is.na(increase))

ny_state_boundaries <- st_read("../tl_2016_36_cousub/tl_2016_36_cousub.shp")
ny_state_boundaries <- st_transform(ny_state_boundaries, crs = 4326)
ny_state_boundaries$COUNTYFP <- as.character(ny_state_boundaries$COUNTYFP)

impact_of_covid_region <- impact_of_covid %>% select(FIPS, County)
impact_of_covid_region <- impact_of_covid_region[!duplicated(impact_of_covid_region), ]
impact_of_covid_region <- merge(impact_of_covid_region, crime_pct_increase, by.x = "County", by.y = "County")
impact_of_covid_region$FIPS <- as.character(impact_of_covid_region$FIPS)

crime_ny_covid_merged <- merge(ny_state_boundaries, impact_of_covid_region, by.x = "COUNTYFP", by.y = "FIPS")

crime_ny_covid_sf <- st_as_sf(crime_ny_covid_merged, coords = c("Longitude", "Latitude"), crs = 4326)

desired_crs <- 4326
library(viridis)

ggplot(crime_ny_covid_sf, aes(fill = increase)) +
  geom_sf(color = "black") +
  scale_fill_viridis_c(option = "magma", direction = -1) +
  theme_minimal() +
  labs(title = "pct increase 2020 -> 2022",
       fill = "increase")

```

Schuyler seems to have a 90% increase since covid:

```{r}
selected_data <- impact_of_covid[impact_of_covid$County == 'Schuyler', ]
selected_data <- selected_data[selected_data$Year > 2015, ]

ggplot(selected_data, aes(x = Year, y = crime_pct, color = County)) +
  geom_line() +
  scale_color_manual(values = my_palette) +
  labs(title = "Crime/population per Year for Counties with least increase post-covid",
       x = "Year",
       y = "Crime/population") +
  theme_minimal()
```


```{r}
impact_of_covid <- crime_ny_with_fips_and_pop
impact_of_covid$crime_pct <- impact_of_covid$Index.Total/impact_of_covid$Population

crime_pct_increase <- impact_of_covid[impact_of_covid$Year >= 2020, ] %>%
  group_by(County) %>%
  summarize(increase = (last(crime_pct) - first(crime_pct))) %>%
  filter(!is.na(increase))

ny_state_boundaries <- st_read("../tl_2016_36_cousub/tl_2016_36_cousub.shp")
ny_state_boundaries <- st_transform(ny_state_boundaries, crs = 4326)
ny_state_boundaries$COUNTYFP <- as.character(ny_state_boundaries$COUNTYFP)

impact_of_covid_region <- impact_of_covid %>% select(FIPS, County)
impact_of_covid_region <- impact_of_covid_region[!duplicated(impact_of_covid_region), ]
impact_of_covid_region <- merge(impact_of_covid_region, crime_pct_increase, by.x = "County", by.y = "County")
impact_of_covid_region$FIPS <- as.character(impact_of_covid_region$FIPS)

crime_ny_covid_merged <- merge(ny_state_boundaries, impact_of_covid_region, by.x = "COUNTYFP", by.y = "FIPS")

crime_ny_covid_sf <- st_as_sf(crime_ny_covid_merged, coords = c("Longitude", "Latitude"), crs = 4326)

desired_crs <- 4326
library(viridis)

ggplot(crime_ny_covid_sf, aes(fill = increase)) +
  geom_sf(color = "black") +
  scale_fill_viridis_c(option = "magma", direction = -1) +
  theme_minimal() +
  labs(title = "absolute increase 2020 -> 2022",
       fill = "increase")
```



## Crime rates (per population size) per counties for 5 year periods


```{r}
combined_data <- crime_ny_filtered_county_tot %>%
  group_by(Year_Period_Label, County) %>%
  summarise(
    Total_Index = sum(Index.Total),
  ) %>%
  ungroup()

top_5_counts <- combined_data %>%
  group_by(Year_Period_Label) %>%
  top_n(5, wt = Total_Index) %>%
  ungroup()

top_5_counts$County <- fct_reorder(top_5_counts$County, 
                                   top_5_counts$Total_Index, 
                                   .desc = TRUE)

top_5_counts <- top_5_counts %>%
  group_by(Year_Period_Label) %>%
  mutate(Rank = -rank(Total_Index))

top_5_counts$County <- fct_reorder(top_5_counts$County, top_5_counts$Rank)


ggplot(top_5_counts, aes(x = County, 
                         y = Total_Index)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Year_Period_Label, 
             scales = "free_x", #"free_x",
             nrow = 5) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Index.Total by County",
       x = "County",
       y = "Index.Total")
```

```{r}

crime_ny_filtered_county_tot <- crime_ny_with_fips_and_pop

crime_ny_filtered_county_tot <- crime_ny_filtered_county_tot %>% select(c('County', 'Year', 'Murder', 'Rape', 'Robbery', 'Aggravated.Assault', 'Burglary', 'Larceny', 'Motor.Vehicle.Theft', 'Population'))

crime_ny_filtered_county_tot <- crime_ny_filtered_county_tot %>% 
  group_by(County, Year) %>%
  summarise_each(list(sum))

crime_ny_filtered_county_tot[ , 3:9] <- 10000*crime_ny_filtered_county_tot[, 2:8]/crime_ny_filtered_county_tot$Population


draw_biplot(crime_ny_filtered_county_tot[crime_ny_filtered_county_tot$Year == 1990 ,c(1,3:9)], project = F)
draw_biplot(crime_ny_filtered_county_tot[crime_ny_filtered_county_tot$Year == 2000 ,c(1,3:9)], project = F)
draw_biplot(crime_ny_filtered_county_tot[crime_ny_filtered_county_tot$Year == 2010 ,c(1,3:9)], project = F)
draw_biplot(crime_ny_filtered_county_tot[crime_ny_filtered_county_tot$Year == 2020 ,c(1,3:9)], project = F)

```

```{r}

crime_ny_filtered_county_tot_2020_to_22 <- crime_ny_with_fips_and_pop[crime_ny_with_fips_and_pop$Year >= 2020,]

crime_ny_filtered_county_tot_2020_to_22_non_ny <- crime_ny_filtered_county_tot_2020_to_22[crime_ny_filtered_county_tot_2020_to_22$Region == 'Non-New York City', ]
crime_ny_filtered_county_tot_2020_to_22_non_ny <- crime_ny_filtered_county_tot_2020_to_22_non_ny %>% select(c('County', 'Murder', 'Rape', 'Robbery', 'Aggravated.Assault', 'Burglary', 'Larceny', 'Motor.Vehicle.Theft', 'Population'))
crime_ny_filtered_county_tot_2020_to_22_non_ny <- crime_ny_filtered_county_tot_2020_to_22_non_ny %>% 
  group_by(County) %>%
  summarise_each(list(sum))

crime_ny_filtered_county_tot_2020_to_22_non_ny[ , 2:8] <- 10000*crime_ny_filtered_county_tot_2020_to_22_non_ny[, 2:8]/crime_ny_filtered_county_tot_2020_to_22_non_ny$Population

draw_biplot(crime_ny_filtered_county_tot_2020_to_22_non_ny[ , 1:8], 'Rape', project = T)
```


```{r}
crime_ny_filtered_county_tot_2020_to_22_ny <- crime_ny_filtered_county_tot_2020_to_22[crime_ny_filtered_county_tot_2020_to_22$Region != 'Non-New York City', ]
crime_ny_filtered_county_tot_2020_to_22_ny <- crime_ny_filtered_county_tot_2020_to_22_ny %>% select(c('County', 'Murder', 'Rape', 'Robbery', 'Aggravated.Assault', 'Burglary', 'Larceny', 'Motor.Vehicle.Theft', 'Population'))
crime_ny_filtered_county_tot_2020_to_22_ny <- crime_ny_filtered_county_tot_2020_to_22_ny %>% 
  group_by(County) %>%
  summarise_each(list(sum))

crime_ny_filtered_county_tot_2020_to_22_ny[ , 2:8] <- 10000*crime_ny_filtered_county_tot_2020_to_22_ny[, 2:8]/crime_ny_filtered_county_tot_2020_to_22_ny$Population

draw_biplot(crime_ny_filtered_county_tot_2020_to_22_ny[ , 1:8], 'Rape', project = T)
```


## Overall crime numbers in NY State and the impact of COVID-10


## Impact of stop-and-frisk on crime statistics under Bloomberg


```{r}
NYC <- crime_ny_with_fips_and_pop[crime_ny_with_fips_and_pop$Region == 'New York City', ]
columns_to_plot <- c("Murder", "Rape", "Robbery", "Aggravated.Assault", "Burglary", "Larceny", "Motor.Vehicle.Theft")
NYC_long <- tidyr::gather(NYC, key = "Variable", value = "Value", columns_to_plot)
NYC_long_summarized <- NYC_long %>%
  group_by(Year, Variable) %>%
  summarise(Sum_Value = sum(Value, na.rm = TRUE),
            Population = sum(Population)) %>%
  ungroup()

# ggplot(NYC, aes(x = Year)) +
#   geom_point(aes(y = Murder), color = "red", linetype = "solid", size = 1) +
#   geom_point(aes(y = Rape), color = "blue", linetype = "solid", size = 1) +
#   geom_point(aes(y = Robbery), color = "green", linetype = "solid", size = 1) +
#   labs(title = "Time Series Plot for Individual Columns in NYC County",
#        x = "Year",
#        y = "Value") +
#   theme_minimal()

ggplot(NYC_long_summarized, aes(x = Year, y = Sum_Value/Population, color = Variable)) +
  geom_line() +
  scale_color_manual(values = my_palette) +
  labs(title = "Crime/population per Year for Counties with least increase post-covid",
       x = "Year",
       y = "Crime/population") +
  theme_minimal()

```

```{r}
non_NYC <- crime_ny_with_fips_and_pop[crime_ny_with_fips_and_pop$Region != 'New York City', ]
columns_to_plot <- c("Murder", "Rape", "Robbery", "Aggravated.Assault", "Burglary", "Larceny", "Motor.Vehicle.Theft")
non_NYC_long <- tidyr::gather(non_NYC, key = "Variable", value = "Value", columns_to_plot)

non_NYC_long_summarized <- non_NYC_long %>%
  group_by(Year, Variable) %>%
  summarise(Sum_Value = sum(Value, na.rm = TRUE),
            Population = sum(Population)) %>%
  ungroup()

# ggplot(NYC, aes(x = Year)) +
#   geom_point(aes(y = Murder), color = "red", linetype = "solid", size = 1) +
#   geom_point(aes(y = Rape), color = "blue", linetype = "solid", size = 1) +
#   geom_point(aes(y = Robbery), color = "green", linetype = "solid", size = 1) +
#   labs(title = "Time Series Plot for Individual Columns in NYC County",
#        x = "Year",
#        y = "Value") +
#   theme_minimal()
# ggplot(non_NYC_long_summarized, aes(x = Year, y = Sum_Value/Population, color = Variable)) +
#   geom_line() +
#   scale_color_manual(values = my_palette) +
#   labs(title = "Crime/population per Year for Counties with least increase post-covid",
#        x = "Year",
#        y = "Crime/population") +
#   theme_minimal()

combined_NYC <- rbind(
  mutate(NYC_long_summarized, Location = "NYC"),
  mutate(non_NYC_long_summarized, Location = "Non-NYC")
)

ggplot(combined_NYC, 
       aes(x = Year, 
           y = Sum_Value/Population, 
           color = Location, 
           group = Location)) +
  geom_line() +
  labs(title = "NYC vs Non-NYC",
       x = "Year",
       y = "Crime/population") +
  scale_color_manual(values = c("NYC" = "blue", "Non-NYC" = "red")) +
  theme_minimal() +
  facet_wrap(~Variable, scales = "free_y")
```

```{r}

combined_NYC_bloomb <- combined_NYC[(combined_NYC$Year > 1999) & (combined_NYC$Year < 2015), ]

ggplot(combined_NYC_bloomb, 
       aes(x = Year, 
           y = Sum_Value/Population, 
           color = Location, 
           group = Location)) +
  geom_line() +
  labs(title = "NYC vs Non-NYC",
       x = "Year",
       y = "Crime/population") +
  scale_color_manual(values = c("NYC" = "blue", 
                                "Non-NYC" = "red")) +
  theme_minimal() +
  facet_wrap(~Variable, scales = "free_y")
```

During 2002 - 2012, NYC imposed stop-and-frisk, we compare NYC crime reports (for 7 major crime stats) vs non-NYC. If the stop and frisk had any effect, we expect NYC crime stats to decrease stronger than non-NYC. Logic: stop and frisk -> "create more safety" -> less crime overall. 

There are various flaws with our analysis: Do not consider other effects, etc.

To compare set starting point the same:

```{r}
combined_NYC_bloomb <- combined_NYC %>%
  filter(Year > 1999 & Year < 2015) %>%
  group_by(Variable, Location) %>%
  mutate(adjusted_value = Sum_Value/Population - first(Sum_Value/Population))

ggplot(combined_NYC_bloomb, aes(x = Year, y = adjusted_value, color = Location, group = Location)) +
  geom_line() +
  labs(title = "Adjusted NYC vs Non-NYC",
       x = "Year",
       y = "Adjusted") +
  scale_color_manual(values = c("NYC" = "blue", "Non-NYC" = "red")) +
  theme_minimal() +
  facet_wrap(~Variable, scales = "free_y")
```


```{r}

combined_NYC_bloomb <- combined_NYC %>%
  filter(Year > 1980 & Year <=2002) %>%
  group_by(Variable, Location)
combined_NYC_bloomb1 <- combined_NYC_bloomb %>%
  group_by(Year, Location) %>%
  summarise(Sum_Value = sum(Sum_Value, na.rm = TRUE),
            Population = sum(Population)) %>%
  ungroup()
combined_NYC_bloomb1 <- combined_NYC_bloomb1 %>%
  group_by(Location) %>% 
  mutate(adjusted_value = Sum_Value/Population - first(Sum_Value/Population))

combined_NYC_bloomb <- combined_NYC %>%
  filter(Year >= 2002 & Year <=2012) %>%
  group_by(Variable, Location)
combined_NYC_bloomb2 <- combined_NYC_bloomb %>%
  group_by(Year, Location) %>%
  summarise(Sum_Value = sum(Sum_Value, na.rm = TRUE),
            Population = sum(Population)) %>%
  ungroup()
combined_NYC_bloomb2 <- combined_NYC_bloomb2 %>%
  group_by(Location) %>% 
  mutate(adjusted_value = Sum_Value/Population - first(Sum_Value/Population))

combined_NYC_bloomb <- combined_NYC %>%
  filter(Year >= 2012 & Year <=2042) %>%
  group_by(Variable, Location)
combined_NYC_bloomb3 <- combined_NYC_bloomb %>%
  group_by(Year, Location) %>%
  summarise(Sum_Value = sum(Sum_Value, na.rm = TRUE),
            Population = sum(Population)) %>%
  ungroup()
combined_NYC_bloomb3 <- combined_NYC_bloomb3 %>%
  group_by(Location) %>% 
  mutate(adjusted_value = Sum_Value/Population - first(Sum_Value/Population))

ggplot(combined_NYC_bloomb1, aes(x = Year, y = adjusted_value, color = Location, group = Location)) +
  geom_line() +
  labs(title = "NYC vs Non-NYC",
       x = "Year",
       y = "Crime/population") +
  scale_color_manual(values = c("NYC" = "blue", "Non-NYC" = "red")) +
  theme_minimal()

ggplot(combined_NYC_bloomb2, aes(x = Year, y = adjusted_value, color = Location, group = Location)) +
  geom_line() +
  labs(title = "NYC vs Non-NYC",
       x = "Year",
       y = "Crime/population") +
  scale_color_manual(values = c("NYC" = "blue", "Non-NYC" = "red")) +
  theme_minimal()

ggplot(combined_NYC_bloomb3, aes(x = Year, y = adjusted_value, color = Location, group = Location)) +
  geom_line() +
  labs(title = "NYC vs Non-NYC",
       x = "Year",
       y = "Crime/population") +
  scale_color_manual(values = c("NYC" = "blue", "Non-NYC" = "red")) +
  theme_minimal()
```


```{r}


ggplot(combined_NYC, aes(x=Year, y=Sum_Value, fill=Variable)) + 
    geom_area()+
  labs(title = "Stacked Area Plot of Crime Types",
       x = "Year",
       y = "Crime")+
  theme_minimal()+
  facet_wrap(~Location, ncol = 1)

```


<!-- 
https://www.nyclu.org/en/stop-and-frisk-data
-->

